Active-contours driven registration of diffusion MR images of the brain

### 2015-03-16 Publications/2015-NeuroImage/05-results.tex
updating paper for second round of reviews
almost ready to revision round 2

% -*- root: 00-main.tex -*-
\section*{Results}
\label{sec:results}

\subsection*{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in section
  \nameref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
To warp the simulated datasets, we randomly generated 150 different realizations of
  the distortion $U_{true}$.
Smoothness is introduced using two levels of B-Spline functions, with control points evenly
  located in isotropic grids covering the full extent of the phantom.
The first level had $50.5mm$ of separation between control points and the second $25.25mm$.
The rationale behind this choice is producing large deformations on the contours with
  the first level, and matching the properties of susceptibility-derived distortions
  as previously reported by \cite{irfanoglu_susceptibility_2011} with the finer grid.
Invertibility of $U_{true}$ is ensured by controlling the maximum displacement at each
  level ($20.2mm$ and $10.1mm$ respectively) as studied in \citep{rueckert_diffeomorphic_2006}.

A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \nameref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below voxel resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel resolution with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

\begin{table}
		\centering
		\footnotesize
    \begin{tabular}{llcc}
    Type      & Low-res.  & Hi-res. \\
    \hline
    ``gyrus'' & 0.58-0.60 & 0.18-0.36 \\
    ``L''     & 0.67-0.73 & 0.35-0.39 \\
    ``ball''  & 0.66-0.73 & 0.32-0.41 \\
    ``box''   & 0.68-0.71 & 0.35-0.39 \\
    \hline
    \end{tabular}
    \caption{All the 95\% confidence intervals (CI) were below the corresponding
      half-voxel resolution in all the phantom types.
    CIs reported in this table were computed using bootstrapping of $10^4$ samples,
      for each phantom type and resolution.
    Taking into account the non-normal distribution of errors, we used the median as location
  		statistic.}\label{tab:ci_phantom}
\end{table}

\subsection*{Correction of distortions on real datasets}\label{sec:results_hcp}

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{}, {\color{red} section S2}),
  we selected the 6-regions scheme described in \nameref{sec:human_connectome}.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure04.pdf}
	\caption{Assessment of the segmentation model, sampling the value of the energy functional
	\eqref{eq:energy} versus several registration errors generated as explained in
	\nameref{sec:res_model_and_metric}.
	The metric of our registration method displayed a smooth gradient towards the minimum
	located at the $\epsilon = 0.0$ point.}\label{fig:energymap}
\end{figure}

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we tested the performance of our registration method, and compared it to a competing
  method: the \emph{b0}-to-\gls*{t2} registration approach.
To this end, we integrated the registration software \emph{elastix} \citep{klein_elastix_2010},
  and reproduced the settings that internally uses \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, a widely used toolkit for tractography analysis of
  \gls*{dti}.
A dual workflow to the general evaluation framework (\autoref{fig:evworkflows})
  was also implemented to run the correction experiment with the competing method.
First, the average \emph{b0} volume is computed using all the \emph{low-b} volumes within
  the distorted \gls*{dmri} dataset.
Using the ``preprocessed'' instance of the \gls*{t2} image of the corresponding subject as
  reference image, and the \emph{b0} image as moving image, we registered both and
  projected the correct (undistorted) contours to the warped space using the resulting
  transform.
For both \emph{regseg} and the competing method, we computed the \gls*{swindex} \eqref{eq:swindex}
  of every surface.
Finally, to statistically compare the results from both methods we performed a one-way ANOVA test
  on the warping indices for three surfaces.
Following the nomenclature introduced in section \nameref{sec:human_connectome}, we selected
  $\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$ as surfaces of interest.
These interfaces are critical in connectivity analyses, and small errors in their location may
  lead to discarding fundamental paths between regions.
We also reported the 95\% CIs of the warping indices for those surfaces.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{The evaluation instrument automatically renders the overlay of the
	  target \gls*{fa} image and the resulting contours after distortion estimation (yellow color) and
	  their theoretical location using the ground-truth (green color).
	The first two rows show several axial slices for \emph{regseg} (``proposed'' row) and the
	  competing method.
	The last two rows represent the sagittal view.
	We intentionally omitted the coronal slicing due to it is the less informative given the properties
	  of the distortion.
	Red arrows point to regions where the accuracy of the proposed method more clearly overperformed
	  the competing method.
	\todo[inline]{This figure will also have the box-plots of the errors of surfaces for both regseg
	and the t2b method}
	}\label{fig:results_real}
\end{figure*}

>>>>

% -*- root: 00-main.tex -*-
\section{Results}
\label{sec:results}

\subsection{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in
  \autoref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \autoref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below the image resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the misregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel size with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).
Since statistical tests may not be conclusive enough, we also computed the confidence intervals
  (\autoref{tab:ci_phantom}).

\begin{table}
		\centering
		\footnotesize
    \begin{tabular}{llcc}
    Type      & Low-res.  & Hi-res. \\
    \hline
    ``gyrus'' & 0.58-0.60 & 0.18-0.36 \\
    ``L''     & 0.67-0.73 & 0.35-0.39 \\
    ``ball''  & 0.66-0.73 & 0.32-0.41 \\
    ``box''   & 0.68-0.71 & 0.35-0.39 \\
    \hline
    \end{tabular}
    \caption{95\% Confidence Intervals (CIs) were computed using bootstrapping of $10^4$ samples,
      for each phantom type and resolution.
    Taking into account the non-normal distribution of errors, we used the median as location
  		statistic.
    All the CIs were below the corresponding half-voxel size in all the phantom types.}\label{tab:ci_phantom}
\end{table}

\subsection{Correction of distortions on real datasets}\label{sec:results_hcp}

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarily, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (\suppl{section S2}),
  we selected the 6-regions scheme described in \autoref{sec:human_connectome}.

@@

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure04.pdf}
	\caption{Assessment of the segmentation model, sampling the value of the energy functional
	\eqref{eq:energy} versus several registration errors generated as explained in
	\autoref{sec:res_model_and_metric}.
	The metric of our registration method displayed a smooth gradient towards the minimum
	located at the $\epsilon = 0.0$ point.}\label{fig:energymap}
\end{figure}

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we tested the performance of our registration method, and compared it to the
  a standard \gls*{t2b} correction.
For both \emph{regseg} and the \gls*{t2b} methods, we computed the \gls*{swindex} \eqref{eq:swindex}
  of every surface.
Finally, to statistically compare the results, we performed a one-way ANOVA test
  on the warping indices for three surfaces.
Following the nomenclature introduced in section \autoref{sec:human_connectome}, we selected
  $\Gamma_{VdGM}$, $\Gamma_{WM}$, and $\Gamma_{pial}$ as surfaces of interest.
These interfaces are critical in connectivity analyses, and small errors in their location may
  lead to discarding fundamental paths between regions.
We also reported the 95\% CIs of the warping indices for those surfaces.
The quantitative results of the ANOVA tests and the CIs are presented in \autoref{tab:results_real}.
Visual assessment of one case is shown in \autoref{fig:results_real}, along with the corresponding
  box plots of error distributions by correction method and surface.
Extended visual assessment of the 16 cases is aggregated in the \suppl{section S4}.

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{The evaluation instrument automatically renders the overlay between the
	  target \gls*{fa} image and the resulting contours after distortion estimation (yellow color) and
	  their theoretical location using the ground-truth (green color).
	The first two rows show several axial slices for \emph{regseg} (``proposed'' row) and the
	  competing method.
	The last two rows represent the sagittal view.
	We intentionally omitted the coronal slicing as it is the least informative, given the directional property
	  of distortions.
	Red arrows point to regions where the accuracy of the proposed method more clearly overperformed
	  the competing method.
	\todo[inline]{This figure will also have the box-plots of the errors of surfaces for both regseg
	and the t2b method}
	}\label{fig:results_real}
\end{figure*}

### 2015-03-16 Publications/2015-NeuroImage/06-discussion.tex
updating intro
almost ready to revision round 2

% -*- root: 00-main.tex -*-
\section*{Discussion}
\label{sec:discussion}
% In section \nameref{sec:methods_background}, we framed our method within the current
%   state of art and defined the contributions in our approach.
% We chose active contours without edges as we were interested in the homogeneous
%   regions of diffusion properties within the brain.
% Iteratively, the active contours are updated following the instant shape gradients
%   (see \nameref{app:shape_priors} in Appendix) with a gradient descent strategy.
% The approach is valid for image processing applications in which a precise segmentation is
%   available (from an atlas or within subject anatomical data), and match any or all
%   the following circumstances:
%   i) reference and target images do not present appropriate contrasts to
%   	perform volume-based registration reliably;
%   ii) the low resolution of the target volume produces strong partial volume effects
%   	that preclude voxel-wise segmentation algorithms and hinder volume-based
%   	registration; and,
%   iii) the estimation of the unknown warping is explicitly needed and the
%     deformation field is smooth enough to be represented with a sufficient number
%   	of B-Spline kernels (i.e. the distortion in \gls*{dmri} images,
%   	object tracking in time series, etc.).
% To support our claim that images can be simultaneously segmented and registered to a
%   reference coordinate system, we conducted evaluation experiments on digital
%   simulated phantoms and real datasets.

@@

\paragraph*{Application on real data}
After a preliminary experiment to demonstrate the aptness of our segmentation model
  in section \nameref{sec:res_model_and_metric}, we evaluated \emph{regseg} in
  a real environment, using the experimental framework presented in \autoref{fig:evworkflows}.
We processed 16 subjects downloaded from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of a competing method.
This alternative approach solves distortion using \emph{b0}-to-\gls*{t2} registration.
Our method presented significantly different results (ANOVA tests are reported in
  \autoref{tab:results_real}), and the 95\% CIs revealed the high accuracy of
  \emph{regseg}.
The aggregated CI for \emph{regseg} was $[0.39 - 0.93] mm$, whereas the competing method
  yielded an aggregated CI of $[2.36 - 3.40] mm$.
Even though we carefully configured the competing method by using the same algorithm and
  extending the same settings from a widely used software, cross-comparison
  experiments are prone to the so-called instrumentation bias \citep{tustison_instrumentation_2013}.
Therefore, these results do not reflect that the \emph{regseg} software \emph{is} better than the
  competing method, but they are communicated to proof that \emph{regseg} is a solid option to
  correct \gls*{dmri} data for distortions, and any other potential application.

\paragraph*{Prospects}
Future extensions or improvements of \emph{regseg} are similar to the evolution of existing
  alternative methodologies.
Incorporating knowledge about the distortion by initializing our method with the theory-based
  correction using fieldmaps would be the most profitable.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{zhou_reliable_2009,daga_susceptibility_2014} to
  circumvent the inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct misregistered areas due to errors in
  phase unwrapping, and underestimations introduced by necessary smoothing
  applied to clean up field maps.
This integration has been recently applied in the \emph{b0}-to-\gls*{t2} approach
  by \cite{irfanoglu_susceptibility_2011}, studied along with the virtues of using
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.

>>>>

% -*- root: 00-main.tex -*-
\section{Discussion}
\label{sec:discussion}

The hypothesis underpinning our research is that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).

@@

\paragraph*{Application on real data}
After a preliminary experiment to demonstrate the aptness of our segmentation model
  in section \nameref{sec:res_model_and_metric}, we evaluated \emph{regseg} in
  a real environment, using the experimental framework presented in \autoref{fig:evworkflows}.
We processed 16 subjects from the \gls*{hcp} database with both \emph{regseg}
  and an in-house replication of a competing method.
This alternative approach solves distortion using \emph{b0}-to-\gls*{t2} registration.
Our method presented significantly different results (ANOVA tests are reported in
  \autoref{tab:results_real}), and the 95\% CIs revealed the high accuracy of
  \emph{regseg}.
The aggregated CI for \emph{regseg} was $[0.39 - 0.93] mm$, whereas the competing method
  yielded an aggregated CI of $[2.36 - 3.40] mm$.
Even though we carefully configured the competing method by using the same algorithm and
  extending the same settings from a widely used software, cross-comparison
  experiments are prone to the so-called instrumentation bias \citep{tustison_instrumentation_2013}.
Therefore, these results do not show that the \emph{regseg} software \emph{is} better than the
  competing method, but they are communicated to proof that \emph{regseg} is a solid option to
  correct \gls*{dmri} data for distortions, and any other potential application.

\paragraph*{Prospects}
Future extensions or improvements of \emph{regseg} are similar to the evolution of existing
  alternative methodologies.
Incorporating knowledge about the distortion by initializing our method with the theory-based
  correction using fieldmaps would be the most profitable.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{zhou_reliable_2009,daga_susceptibility_2014} to
  circumvent the inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct misregistered areas due to errors in
  phase unwrapping, and underestimations introduced by necessary smoothing
  applied to clean up field maps.
This integration has been recently applied in the \emph{b0}-to-\gls*{t2} approach
  by \cite{irfanoglu_susceptibility_2011}, also investigating the usefulness of
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.

### 2015-03-16 Publications/2015-NeuroImage/07-conclusion.tex
almost ready to revision round 2

\emph{Regseg} is a variational framework to simultaneously segment and
  register 3D \gls*{dmri} data of the human brain, using within-subject
  anatomical information as reference.
The registration method segments the target multivariate image in several competing regions
  defined explicitly by their limiting surfaces.
The surfaces are active, and evolve on a B-Spline displacements field, guided by shape
  gradients computed on the current partition of the target image.
\emph{Regseg} uses active contours without edges, and hence looks for
  homogeneous regions within the image.
We tested \emph{regseg} on digital phantoms simulating \gls*{t1} and \gls*{t2} \gls*{mri},
	warped with smooth random deformations.
The resulting average misplacement of the contours was significantly lower than the
  image resolution of the phantoms.

>>>>

\emph{Regseg} is a variational framework to simultaneously segment and
  register 3D \gls*{dmri} data of the human brain, using within-subject
  anatomical information as reference.
The registration method segments the target multivariate image in several competing regions
  defined explicitly by their limiting surfaces.
The surfaces are active, and evolve on a free-deformation field supported by B-Splines.
A descent optimization strategy is guided by shape gradients computed on the current partition
  of the target image.
\emph{Regseg} uses active contours without edges, and hence looks for
  homogeneous regions within the image.
We tested \emph{regseg} on digital phantoms simulating \gls*{t1} and \gls*{t2} \gls*{mri},
	warped with smooth random deformations.
The resulting average misplacement of the contours was significantly lower than the
  image resolution of the phantoms.
